\section*{Measures of Central Tendency}

\textbf{Desirable properties} of a satisfactory average (or measure of dispersion) according to G. U. Yule:

\begin{itemize}
    \item It should be rigidly defined.
    \item It should be based on all observations.
    \item It should be readily comprehensible.
    \item It should be easily calculable.    
    \item It should be affected as little as possible by sampling fluctuations.
    \item It should readily lend itself to algebraic treatment.
\end{itemize}

\subsection*{Mean}
\textbf{Arithmetic Mean (AM)}

\begin{itemize}
    \item Simple A.M.: $ \bar{x} = \frac{1}{n}\sum_1^n x_i $
    \item $x_i$: observations, $n$: number of observations.
    \item Weighted A.M.: $ \bar{x} = \frac{1}{N}\sum_1^n f_i x_i $
    \item $f_i$: frequencies, $N$: $\sum_1^n f_i$.
\end{itemize}

\textbf{Geometric Mean (GM)}

\begin{itemize}
    \item Simple G.M.: $ g = \sqrt[n]{x_1 x_2 \dots x_n}$
    \item Weighted G.M.: $ g = \sqrt[N]{x_1^{f_1} x_2^{f_2} \dots x_n^{f_n}} $
\end{itemize}

\textbf{Harmonic Mean (HM)}
\begin{itemize}
    \item Simple HM: $ h = \frac{n}{\sum_i \frac{1}{x_i}} $
    \item Weighted HM: $ h = \frac{N}{\sum_i \frac{f_i}{x_i}} $
\end{itemize}

\subsection*{Median and Mode}
\textbf{Median}
\begin{itemize}
    \item Odd $n$: $ M = x_{(n+1)/2} $
    \item Even $n$: $ M = \frac{x_{n/2} + x_{n/2+1}}{2} $
\end{itemize}

\textbf{Mode}
\begin{itemize}
    \item The most frequently occurring value.
\end{itemize}

\subsection*{For Grouped Data}
\textbf{Mean:} $ \bar{x} = \frac{\sum f_i x_i}{\sum f_i} $

\textbf{Median:} $ l_1 + \frac{\frac{N}{2} - F}{f_m} \times c $
\begin{itemize}
    \item $l_1$: lower boundary of median class
    \item $N$: total frequency
    \item $F$: cumulative frequency below $l_1$
    \item $f_m$: frequency of median class
    \item $c$: width of median class
\end{itemize}

\textbf{Mode:} $ l_1 + \frac{d_1}{d_1 + d_2} \times c $
\begin{itemize}
    \item $l_1$: lower boundary of modal class
    \item $d_1$: difference between frequency of modal class and preceding class $(f_1 - f_0)$
    \item $d_2$: difference between frequency of modal class and following class $(f_1 - f_2)$
    \item $c$: common width of classes
\end{itemize}

\subsection*{Properties of AM}
\begin{itemize}
    \item $ \sum x_i = n \bar{x} $ (Total of observations)
    \item $ \sum (x_i - \bar{x}) = 0 $ (Sum of deviations)
    \item If $ z = ax + b $, then $ \bar{z} = a \bar{x} + b $
    \item Mean of combined groups:
    \[ \bar{x} = \frac{n_1 \bar{x}_1 + n_2 \bar{x}_2}{n_1 + n_2} \]
    \item $ \sum (x_i - A)^2 $ is minimized when $ A = $ simple A.M.
    \item $ \sum f_i(x_i - A)^2 $ is minimized when $ A = $ weighted A.M.
\end{itemize}

\subsection*{Properties of GM}
\begin{itemize}
    \item $ x_1 x_2 \dots x_n = g^n $
    \item $ \ln g = \frac{1}{n} \sum \ln x_i $
\end{itemize}

\subsection*{Relations between AM, GM, HM}
\begin{itemize}
    \item $ AM \geq GM \geq HM $ (Equality holds if all values are equal.)
    \item For two values, $ \frac{AM}{GM} = \frac{GM}{HM} $
\end{itemize}

\subsection*{Mean, Median, Mode Relationship}
\begin{itemize}
    \item $ \text{Mean} - \text{Mode} = 3(\text{Mean} - \text{Median}) $
\end{itemize}

\subsection*{Partition Values}
\textbf{Quartiles:}
\begin{itemize}
    \item $ Q_1 < Q_2 < Q_3 $
    \item Median = $ Q_2 $
    \item Quartile deviation: $ \frac{Q_3 - Q_1}{2} $
    \item Skewness: $ \frac{Q_3 - 2Q_2 + Q_1}{Q_3 - Q_1} $
\end{itemize}

\textbf{Deciles:}
\begin{itemize}
    \item $ D_1 < D_2 < \dots < D_9 $, $ D_5 = Q_2 $
\end{itemize}

\textbf{Percentiles:}
\begin{itemize}
    \item $ P_1 < P_2 < \dots < P_{99} $
    \item $ P_{10} = D_1, P_{20} = D_2, \dots, P_{90} = D_9 $
    \item $ P_{25} = Q_1, P_{50} = Q_2, P_{75} = Q_3 $
\end{itemize}

